\tit

### 2021-05-31 response-to-reviews.md

# Paper #87 review responses

***Note:** I say "we" instead of "I" in these responses, because the blind review process was still in effect when I wrote them.*

> The authors’ response should address reviewer concerns, answer reviewer questions, and correct misunderstandings. Reviewers are not required to read past the 500 word limit, so please focus your response there. You may provide additional (e.g. point-by-point, per-reviewer) response beyond the limit if it may help the reviewers.

***Oops. I read this clarification late in the game. To do: reorganize the following remarks accordingly.***

## Review A

> My major concern is novelty, but unfortunately I am not familiar with the prior work. Given that I was a bit surprised by the absence of prior work, and that there were already so many papers on related topics, my rating is B. I might advocate for this paper more enthusiastically once novelty is confirmed.

We were surprised as well not to find formalized & mechanized connections between the simple, non-operational semantics of language specification and the dual decidable representations of regular expressions and tries.
Note also that proof relevance yields *parsing* (explanations) rather than mere recognition.
We are not aware of formalized (or even rigorous) accounts of *parsing* (as opposed to mere recognition) based on non-operational specification of languages and derivatives.

> I recognize that a note on the versions of Agda and its standard library would help the reader.

Will do. Thank you for the suggestion!

## Review B

> It seems that in Fig. 1 "a language over an alphabet A" is not A^* but A^* -> Set l. This is at least nonstandard --- a standard definition should be A*. Explain the reason.

`A^*` is the set of *all* strings over an alphabet `A`, i.e., the "universal language".
Other languages are proper subsets, which are neatly formalized in type theory as `A^* -> Set l`, i.e., a function mapping each string `w` to the type of proofs that `w` is in the language.

> By the way, can you use Prop instead of Set l? I suspect the former is more intuitive, especially for those who are not experts on type theory.

`Set` appears to be much more widely used in the Agda ecosystem than `Prop` and so using `Set` allows reusing many more existing definitions and proofs, including type isomorphisms.
More fundamentally, `Prop` is proof-irrelevant, so we would only get language recognition.
In contrast, using `Set` gives proof-relevant recognition, i.e., parsing.

> In Fig. 1, while I believe c :: A, I could not figure out the type of scalar, s.

In the definition of scalar multiplication (`(s · P) w = s × P w`), `s` appears as an argument to the type product operator and so is a type (i.e., has type `Set ℓ`).
Logically, it's a proposition being conjoined with the property of the string `w` being in the language `P`.
This operator makes for more elegant and more general formulations of derivatives of concatenated and starred languages than Brzozowski's definitions.
The additional generality comes from proof relevance ("parsing").
It's also fundamentally important to the algebraic setting of semimodules (generalized vector spaces), which is a natural fit to proof-relevant languages.

> This paper carefully distinguished three kinds of equalities, namely propositional equality, isomorphism, and extensional isomorphism. However, I did not understand why that equality should be used for each occurrence. For example, why in Fig. 1 is propositional equality necessary?

Figure 1 uses both propositional equality and extensional isomorphism.
We could not have used (propositional) equality for all of the lemmas, since this strong condition does not hold in all cases.
We could have used the weaker isomorphism condition for all of the lemmas, but the equalities are more specific and thus more informative.
Moreover, isomorphisms require some notational overhead to apply, as shown where they are invoked in figures 5 and 7.
In fact, there might be some notational overhead even for propositional equalities, but these ones are actually *definitional* (meaning proved by normalization, as noted on line 194), making them especially convenient.

The reason for using both isomorphism and extensional isomorphism is simply that the `ν` lemmas are about types/propositions (nullability) while the `δ` lemmas are about predicates (general, proof-relevant language membership).

*Correction:* My answer was about Figure 3.
As for Figure 1, the uses of propositional equality are for lists (not types or type-level predicates), to which the other mentioned equality notions do not apply.

> Section 2 uses A* -> B instead of A* -> Set l. Perhaps, the reason is that all the definitions can be polymorphic on the result, B. However, this difference at least makes the meaning difficult to understand.

Indeed, this generalization is important because it points out that the central notion of derivatives is much more general than its usual setting of languages (even beyond generalizing from `Set` to arbitrary semirings), and some definitions and properties hold in the more general setting.

> p5. "isomorphic types or predicates are equivalently decidable":
> I could not understand the reason.

The reason (proof) is outlined in the next sentence ("One direction of the isomorphism proves \AB{B} from \AB{A}, while the other proves {\AF ¬ B} from {\AF ¬ A}") and formalized in the definitions/proofs of the `_▹_` and `_▸_` signatures/lemmas stated in between.
As with the rest of the paper, we mostly state the lemmas & theorems (signatures), leaving the interested reader to view proofs as desired in the paper/project's Agda source code.

> Figs. 4 and 5 are nearly identical to Figs. 6 and 7. While I could understand that it is because of the duality, I still wonder what is the essential differences.

The essential difference is that the representation data constructors and the representation examiners trade places (the duality).
More specifically

*   For regular expressions, language building blocks (`∅`, `_∪_`, `𝟙`, `_*_`, etc) are data constructors, while `ν` and `δ` are functions that examine the data.
*   For tries, `ν` and `δ` are codata constructors, while language building blocks are functions that examine the codata.

This duality is visually highlighted in the code arrangement in figures 4 and 6 and in the syntax coloring in figures 5 and 7.
(Line 419: "The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.")

> Usually, an inductive definition gives a datatype, whereas a coinductive one does a function (or stream). However, the automatic differentiation requires the concrete construction process of the function even for the coinductive case. Another possible difference is the potential of the infiniteness of the coinductive definition. However, for this particular case, I am not aware of the usefulness of infiniteness.

The same pattern occurs in this paper.
Streams are a special case of string tries (for a singleton alphabet), being infinite in the specific and more general case.
The pattern also generalizes beyond string tries to tries over arbitrary regular algebraic data types.

By "concrete construction process", do you mean that there are still (corecursively defined) examination functions?
If so, this situation is also customary in our experience when working with coinductive structures.

> On page 9. it is argued that a difference is a work-sharing between the function and its derivative. I agree with this point but do not think it to be critical. Existing parsing methods based on the derivatives do such work sharing. Moreover, automatic differentiation alone cannot share repeated calculations of derivatives when we arrive at the same state more than once.

We thought this work sharing issue worth pointing out, because it makes an enormous difference for symbolic vs automatic differentiation in the usual calculus sense, and indeed is the main argument usually given in the literature for preferring the latter over the former.
We are not aware of a clear or rigorous discussion of the issue in derivative-based parsing work (though there are many optimizations described informally and without semantic basis).
It is not clear what the practical implications are in the setting of languages, but it seems worth mentioning to stimulate closer examination.
If the performance difference is much less than with differential calculus, then it's important to understand why.
For instance, maybe there is a dual technique for regular expressions that could be examined clearly for useful practical and theoretical insight.
On the other hand, if the difference is the same or greater, then it's important to exploit the potential with better justification and perhaps implementation.

We certainly agree with your remark about the potential for improved work sharing when a "state" is reached in more than one way!

## Review C

> The specification of language membership (language operations) is non-operational and it seems to be one of the key ideas that the authors are advocating, i.e., to have as simple as possible specifications with respect to which other operations (or implementations) can be proved correct.

Yes, exactly.

> The result of this approach is that the provided language membership deciders (defined using differentiation on inductive or coinductive representations of the language) are correct. In other words, the function that decides language membership is a theorem saying that if there is a representation for a language P, then language membership in P is decidable, i.e., for any w, the function either computes a witness of P w (the type of language membership proofs of w in P) or constructs a proof that no such exists.

Indeed. Well-summarized. Thank you.

> Overall, I think this is a nice, clean, well-written and compact paper. Working things out so that clean and simple definitions are also "magically" correct according to Agda is often a non-trivial task. The idea that specifications should be as simple as possible is a very reasonable one. Therefore, I think these ideas deserve to be communicated.

Thank you!
The clean, simple, automatically verified nature is exactly what we were wanting to highlight.

> One issue I have is that reading the paper might give the impression that the choice of type theory (instead of set theory) or type-level languages is necessary for achieving non-operational specification of language membership or more informative language membership than just true/false. >The paper does demonstrate that type theory can be convenient for achieving this.
>
> For example, the concatenation of P and Q (subsets of A*) is a certain subset of A* 
>
> P * Q = { w \in A* | exists u, v. w = uv \and u \in P \and v \in Q }
>
> and the characteristic function of this set corresponds to what is given in Figure 1. This could be done in Agda with decidable languages (A* -> 2) and thus also achieve non-operational specification (values of A* -> 2 as type indices). (Assumption: a word w can be decomposed in a finite number of ways.)

We wrote a comparison with set-theoretical and decidable specifications and discarded it for better focus.
Yes, one can start with decidable languages and exploit the property that every finite list can be unappended in finitely many ways.
Doing so would go against the principle of simple, elegant, non-operational *specifications*, since it relies on cleverness involving and motivated by decidability of language *recognition* rather than the simpler nature of language *definition*.

> line 18: I think it would be helpful to later mention what properties would have required induction/coinduction?

Thanks. We might refer to Abel [2016], which proves a great number of algebraic properties on tries, all of them needing coinduction/bisimulation.
The key to eliminating these much more complex proofs is inheriting properties from the *indices* (languages specified as type-level predicates), which are simple and neither inductive nor coinductive.

> Contribution 1, line 93: Is this meant to say that capturing the semantic essence of Brzozowski's derivatives is something new? The semantic operation was defined before the syntactic one. Brzozowski references three papers where (under different names) derivatives of sequences are used and then defines the corresponding operation on syntax.

*Formalizing* (stating and proving in a formal system) the semantic essence is new, as far as we know. Generalizing this essence from languages to functions from lists might be new as well. Moreover, most recent work we know of seems to identify derivatives with regular expressions, leading readers away from the more natural and general notion in practice and its alternative realizations, such as tries. (Abel [2016] is an exception but lacks a formal semantic specification separate from the implementation itself.)

See also the paragraph starting on line 717, including the mention of Brzozowski on line 720.
I imagine you will agree that it is easier to make clear comparisons after the technical substance is presented (in conclusions) than before (in introduction); and yet readers want at least a general sense of comparison before the technical substance.

> Contribution 2, line 99: The duality is between regular expressions and a certain subset of tries (those constructed using the language building operations). The two things named Lang in Figures 4 and 6 are not equally expressive.

We agree. This comparison can be expressed more clearly, including the fact that regular expressions *and their derivatives* also naturally generalize to a much broader class of languages, as Might and Darais [2010] and follow-up work has noted (though without much specification or rigor).

> line 103: Is "proof isomorphism" (throughout the paper) the same as "type isomorphism" on line 188?

Yes. "Type isomorphism" is more technically specific (suiting the technical meat of the paper), while its use as "proof isomorphism" seems more suitable for the introduction and the contrast with "logical equivalence" mentioned in line 103.

> Contribution 3, line 103: Is this statement considering previous work on formalisations or formal languages in general? As soon as something more general is considered instead of nullability (true/false empty word property), the lemmas (or definitions) for concatenation and closure start to look like those in Figure 2.

The statement refers to work we are aware of in formal languages as well as its formalization. Have we missed something important?

> line 179: What is meant exactly by "liberation"? I think it is fairly standard to consider left and right quotients (wrt. a word) of languages (sets of strings, no grammars) as monoid actions and thus reducing language membership to nullability also in a step-by-step manner.

This sentence is in the context of the immediately preceding sentence, which refers to Brzozowski's paper ("Derivatives of *Regular Expressions*"---emphasis mine) and its extensions by Might and Darais [2010] etc.
We are trying to encourage a shift from recent near-identification of derivatives (for parsing) as about a particular grammatical/inductive language representation back to the underlying semantic basis you mention (though also tilted toward proof relevance for parsing and other applications), so that the formal development has a simpler formal starting point (and neither inductive nor coinductive) and alternatives to grammatical representation are easier to notice.

> line 283: I think it would be good to also say what is the correctness condition (in addition to the fact that it is guaranteed by types).

The condition is stated on line 228 and is captured in the type of `⟦_⟧` on line 310, namely that `⟦_⟧` is both computable and converts a language into a decidable recognizer/parser.
We will add a reminder at line 283.
Thanks.

> line 308: This left triangle looks like a map operation. Maybe a comment about why this has to be part of the syntax.

The left triangle *reifies* a map operation.
The reason it must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (figure 2) translate into an implementation.
We will add a comment.

> line 333: Maybe a comment about why f works on the left hand side of the triangle with both p and with p differentiated.

Thanks.
The reason is that both `p` and `δ p a` are languages/predicates.

> Figures 4 and 5: there is no definition of the language (A* -> Set) denoted by an expression and thus no formal statement that it indeed denotes the language in its type.

Thank you.
This missing semantic function maps every `p : Lang P` to `P`, regardless of the specifics of `p` (which is why the algebraic properties transfer effortlessly from type-level languages without having to consider the structure of `p`).
We agree that it's worth stating explicitly and will do so.

> line 623: By a choice of an equivalence relation we are quotienting the (encapsulated) representation. It would be good to give a bit more intuition here (which quotient is this).

Indeed. This comment is closely related to your remark above about the denotation of the decidable language representations and has the same answer. Equivalence of representations is *semantic* equivalence, which is to say equivalence of the indices (type-level languages) that capture the semantics.
We will add this clarification.

> line 723: What are transformations?

The rules of differentiation, as transformations of symbolic/grammatical representations.

> line 724: "based *on* the"?

Fixed. Thanks!

## Review D

> Section 6.2 is quite dense, and its motivation is not entirely clear to me. Is its purpose to show that existentials can be used to force the dependently-typed definitions developed throughout the paper into the simply-typed algebraic structures? If so, I believe a single paragraph would to, and would perhaps be less confusing. Furthermore, the section title is confusing: transport properties to what? We haven't defined the existentially quantifier variants yet, so it's not yet obvious that we have anywhere to transport properties to.

The purpose is to show that the properties of non-decidable, type-level languages transfer effortlessly to the decidable implementations.
The importance of this fact is that one gets both (a) easy proofs of useful correctness properties (without complications due to efficiency and even decidability) *and* (b) efficient implementations.

This simple solution for property transfer, however, raises a "technical obstacle" (line 581): the indexed data types do not match the types of standard algebraic properties.
Existential wrapping is a simple, generic solution to this obstacle.

I renamed the section from "Transporting Properties" to "Transporting Properties from Specification to Implementations"

> Line 86: The clause "including the use of existential quantifiers" is confusing, as it in no way makes clear what the existential quantifiers are used for. I'd suggest dropping the clause entirely.

I've appended "(as in Figure 1) to that sentence".

> Line 500: The types of the primitive language constructs are elided throughout the paper. Therefore, it is a bit surprising to see that the variant of star which, ostensibly, abstracts away from lists, still references the list primitives foldr and All, as well as the monoid properties.

Before generalizing, the list plays two different algebraic roles: monoid and foldable.
Either can be generalized independently.

> Line 608: Why are the existentially-wrapped versions specifically based on the inductive variant? Is there anything specific that makes these better or more amenable? The text so far led me to believe the coinductive variant was better and more performant.

It was an arbitrary choice. Both implementations (inductive and coinductive) are wrapped alike. We will say so in the text.

> Line 638: Is the pattern matching in the type valid Agda syntax? Again, I don't think these liftings really add anything to the paper.

Yes. All such declarations and definitions in the paper (and those elided from the paper) are fully type- (and thus correctness-)checked when the paper is typeset.

* * * * * * * * * * * * * * * * * * * *

From the final review:
<blockquote>
The reviewers want to thank the author for this interesting submission and are happy to accept it with the understanding that the author will make all of the following changes:

1. Clarify what's new in the context of proof mechanization or type theory, what's new in the context of parsing, and what's new in the context of formal languages & automata. In particular, some contributions do not seem to be novel in the last context. The author did not respond to Reviewer C's comment about the work of Lombardy and Sakarovitch, but their work seems to be similar to the derivative operations in the paper and the author should pay attention to it (in addition to other related works).

2. Explain the Agda syntax more for the readers who are not familiar with Agda. For example, add concrete examples showing how operators in Fig. 1 apply to concrete regular expressions and words, including the fully expanded types and values (as witnesses of the membership). The current draft has only 16 pages so space should not be an issue.

3. Clarify the difference between data and codata: both generally and in terms of difference it makes for main results of this paper.

4. Consider explaining how to recover the ordinary setting where a language is a set of words and language union is idempotent. For example, explain how the current work can be used to prove the classical theorem for the ordinary setting that the set of derivatives of a regular expression is finite modulo associativity, commutativity and idempotence of union (although the derivative defined here might require more).

5. Fix typos and other issues that have already been discussed in the reviews and the response.
</blockquote>

*My response:*

1.  I added the following paragraph to the beginning of section 7 (Related Work):

    > The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
    > In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.

    Also an initial sentence to the (now) third paragraph of that section:

    > The main contributions of this paper are to formalization and mechanization of language recognition and parsing.

2.  I added a new figure 2 with examples of languages and membership proofs, along with a new footnote 2 giving some detailed explanations.

3.  I added a new footnote 10.

4.  Appended to the final paragraph of section 7:

    > As mentioned at the end of \secref{Decidability}, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence.
    > Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus replacing parsing by recognition.
    > One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).

5.  Done!

### 2021-05-31 paper.tex
processed review comments & requirements

\ifacm
\documentclass[acmsmall,screen,timestamp,anonymous,review]{acmart}
\else
\documentclass[acmsmall,screen,timestamp,nonacm]{acmart}
\fi
\authorsaddresses{}

@@

\item The types {\AB A \AD ⊎ \AB B} and {\AB A \ARe × \AB B} represent disjunction and conjunction respectively as non-dependent sum and products.

@@

Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} \AF ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
(If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
For instance, every proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}} \emph{and} the knowledge of which one was chosen (with different proofs reflecting different choices).
Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}}\out{ (language concatenation)} includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{w} \AD ≡ \AB{u} \AF ⊙ \AB{v}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.

The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
As described in \secref{Properties}, it is also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis for specification and reasoning, with correct and computable parsing implementations following systematically as corollaries.

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated and proved about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions (\secref{Decomposing Languages}).
\item
  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secref{From Languages to Parsing}).
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than the usual, coarser relation of logical equivalence.
  As such, the lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings distinctly, including closure of languages containing the empty string.
\item
  The vocabulary of languages is generalized from predicates on lists to predicates on arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
  This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).
  % Properties are proved in terms of this more general vocabulary.
\item
  Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties}) with no additional proofs needed.
\end{itemize}
Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.%
\ifacm\else
%% \citep{Elliott2021-derivatives-repo}.
%% (\repoURL).
\footnote{\protect\repoURL}
\fi{}
All type signatures and definitions in the paper are generated by the Agda compiler.

@@

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} for \AF{𝒟foldl}):
\agda{foldl}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), as will be exploited in \secref{Automatic Differentiation}.

@@

\item \AF{inj₁} and \AF{inj₂} are left and right injections into a sum type.

@@

\item \AF{\_✶‽} reflects the fact that the empty list exists for every element type (even \AF ⊥).

}
Moreover, decidability lifts naturally from types to predicates:
\agda{Decidable}
With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.

@@

The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, while the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
An automatic solution would need some form of \emph{reflection} of language-constructing operations or of \AF{ν} and \AF{δ} as inspectable data.

@@

\secreftwo{Symbolic Differentiation}{Automatic Differentiation} applies the symbolic and automatic strategies respectively to languages.
Both algorithms are corollaries of the lemmas in \figref{nu-delta-lemmas} and are automatically proved correct by type checking.

@@

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, we can use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a function {\AF{⟦\_⟧}} that converts a symbolic language representation to a computable parser.

Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These clauses are meant to be read in ``column-major'' order, i.e., each column is a complete definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
(We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.)

>>>>

\ifacm
\documentclass[acmsmall,screen,timestamp]{acmart}  % ,anonymous,review
\else
\documentclass[acmsmall,screen,timestamp,nonacm]{acmart}
\fi
\authorsaddresses{}

@@

\item For types (propositions) \AB A and \AB B, the types {\AB A \AD ⊎ \AB B} and {\AB A \ARe × \AB B} represent disjunction and conjunction respectively as non-dependent sum and products.

@@

Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} \AF ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
(If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
For instance, every proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}} \emph{and} the knowledge of which one was chosen (with different proofs reflecting different choices).
Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}}\out{ (language concatenation)} includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{w} \AD ≡ \AB{u} \AF ⊙ \AB{v}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
\rnc\source{Examples}
\figrefdef{language-examples}{Language examples}{\agda{examples}} contains some examples of languages (left) and corresponding membership proofs (right).%
\footnote{Some explanatory comments:
\begin{itemize}

\item Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{a∪b☆}''.

\item The alphabet type (\AB A in \figref{Lang-ops}) used for these examples is \AF{Char}, i.e., characters.

\item Single-item lists are notated with square brackets.

\item Variables defined but not used can be replaced by the anonymous pattern ``\AK{\_}'' (as in proofs on the right).

\item The \AF{inj₁} and \AF{inj₂} functions are left and right injections into a sum type, corresponding to the left and right introduction forms for logical disjunction.

%% \item The \AF{inj₁} and \AF{inj₂} functions are left and right injections/introductions into a sum/disjunction type.

\item Pairing\out{ (indicated by the infix comma operator)} is right-associative.
      Since the second and third examples are existential propositions, their inhabiting proofs are dependent pairs.
      Each comprises a division into substrings and a proof that the substrings concatenate to the given test string (the language's argument), along with language membership proofs for each substring.

\item Equality proofs in these examples are by normalization and hence given by \AIC{refl}.
\vspace{-2ex}% Avoid bogus blank footnote line spilling to next page
\end{itemize}}

The use of type-level predicates makes for simple, direct specification, including the use of existential quantification (as in \figref{Lang-ops}).
As described in \secref{Properties}, it is also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis for specification and reasoning, with correct and computable parsing implementations following systematically as corollaries.

This paper makes the following specific contributions:
\begin{itemize}
\item
  A collection of lemmas are stated and proved about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions (\secref{Decomposing Languages}).
\item
  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secref{From Languages to Parsing}).
  Each implementation contains exactly the same code but in transposed forms.
  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
\item
  The theory and algorithms are specified in terms of proof isomorphism rather than the usual, coarser relation of logical equivalence.
  As such, the lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings distinctly, including closure of languages containing the empty string.
\item
  The vocabulary of languages is generalized from predicates on lists to predicates on arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
  This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).
  % Properties are proved in terms of this more general vocabulary.
\item
  Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties from Specification to Implementations}) with no additional proofs needed.
\end{itemize}
Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.%
\ifacm\else
%% \citep{Elliott2021-derivatives-repo}.
%% (\repoURL).
\footnote{\protect\repoURL}
\fi{}
All type signatures and definitions in the paper are generated by the Agda compiler.

@@

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Recall that Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} for \AF{𝒟foldl}):
\agda{foldl}
\vspace{-5ex}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), as will be exploited in \secref{Automatic Differentiation}.

@@

\item Recall that \AF{inj₁} and \AF{inj₂} are left and right injections into a sum type.

@@

\item \AF{\_✶‽} reflects the fact that the empty list exists for every element type (even \AF ⊥).
\end{itemize}
\vspace{-2ex}
}
Moreover, decidability lifts naturally from types to predicates:
\agda{Decidable}
With these definitions, we can succinctly formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.

@@

The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, while the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to structural inspection.
An automatic solution would need some form of \emph{reflection} of language-constructing operations or of \AF{ν} and \AF{δ} as inspectable data.

@@

\secreftwo{Symbolic Differentiation}{Automatic Differentiation} apply the symbolic and automatic strategies respectively to languages.
Both algorithms are corollaries of the lemmas in \figref{nu-delta-lemmas} and are automatically proved correct by type checking.

@@

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivial semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
The reason \AF{\_◂\_} must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (\figref{nu-delta-lemmas}) translate into an implementation in terms of that representation.

Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧‽}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
These clauses are meant to be read in ``column-major'' order, i.e., each column is a complete definition.
Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
(Recall from \secref{Decidability} that correctness is defined as constructing a term of type {\AF{Decidable} \AF{P}} for a given language \AB{P}, as satisfied by the type of {\AF{⟦\_⟧‽}}.)
We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.
